Comparing SAT and ILP formulations for Computational and Systems Biology problems

### 2020-11-10 paper.tex
hipp-sat figure
updates

Denote by $\mathcal{G}=\{g_i\mid 1\leq i \leq n\}$ a set of $n$ genotype vectors of length $m$ and by $g_{ij}$ the $j$'th entry of genotype vector $g_i$.
We construct a propositional formula $\phi (\mathcal{G}, r)$ which is satisfiable if and only if there exists a set of haplotypes $\mathcal{H}$ of size $r$ such that each genotype $g_i\in \mathcal{G}$ is explained by some pair fo haplotypes $(h_j, h_k)$ in $\mathcal{H}$.
To solve the HIPP problem the satisfiability of $\phi (\mathcal{G}, r)$ is checked for increasing values of $r$ until a value is found such that $\phi (\mathcal{G}, r)$ is satisfiable but $\phi (\mathcal{G}, r-1)$ is not.
This value of $r$ is then the minimum number of haplotypes required, and the actual haplotypes can be extracted from the satisfying assignment to $\phi (\mathcal{G}, r)$.

The formula $\phi (\mathcal{G}, r)$ considers a set of $r$ distinct haplotypes $h_k$ and encodes the assignment of a pair $(h_{k_1}, h_{k_2})$ to each $g_i$ using Boolean variables $s^a_{k_1i}$, $s^b_{k_2i}$ and $h_{kj}$.
That is, variables $s^a_{k_1i}$ and $s^b_{k_2i}$ being set true indicates that $(h_{k_1}, h_{k_2})$ is the pair of haplotypes assigned to genotype $g_i$.
The variable $h_{kj}$ is the $j$'th entry of haplotype $h_k$.
The formula then requires two types of constraints
\begin{enumerate}
\item[(i)] The selector variable $s^a_{ki}$ should select a unique haplotype, i.e., for each genotype $g_i$ exactly one variable $s^a_{ki}$ should be set true (and similarly for $s^b_{ki}$). 
\item[(ii)] Each genotype $g_i$ should be explained by its selected pair of haplotypes $(h_{k_1}, h_{k_2})$. 
\end{enumerate}

Constraint (i) corresponds to equations $\sum_{1\leq k\leq r} s^a_{ki} = 1$ and $\sum_{1\leq k\leq r} s^b_{ki} = 1$ holding for all $i$ such that $1\leq i \leq n$.
This constraint is enforced in $\phi (\mathcal{R}, r)$ utilizing a propositional encoding of an \emph{1-bit adder circuit}.
The adder circuit uses variables $v^a_{ki}$ where $1\leq k \leq r+1$ and $1\leq i\leq n$ and the following constraints (the ones for $s^b_{ki}$ are similar).
\begin{eqnarray}
\forall i\text{ such that } 1\leq i\leq n: & v^a_{1i} \leftrightarrow s^a_{1i}\label{hipp-sat:clause1}\\
\forall k, i\text{ such that } 1\leq k \leq r,\ 1\leq i\leq n: & \neg v^a_{ki} \vee \neg s^a_{ki}\label{hipp-sat:clause2}\\
\forall k, i\text{ such that } 1\leq k \leq r,\ 1\leq i\leq n: & v^a_{(k+1)i} \leftrightarrow (s^a_{ki}\vee v^a_{ki} )\label{hipp-sat:clause3}\\
\forall i\text{ such that } 1\leq i\leq n: & v^a_{(r+1)i}=1\label{hipp-sat:clause4}
\end{eqnarray}
Constraint \eqref{hipp-sat:clause1} enforces $v^a_{1i}$ and $s^a_{1i}$ to take on the same value in every satisfying assignment whereas constraint \eqref{hipp-sat:clause4} enforces variable $v^a_{(r+1)i}$ to be evaluated to true.
Constraint \eqref{hipp-sat:clause2} prevents $v^a_{ki}$ and $s^a_{ki}$ from being simultaneously assigned true, 
and constraint \eqref{hipp-sat:clause3} sets $v^a_{(k+1)i}$ to 1 whenever either $s^a_{ki}$ or $v^a_{ki}$ is assigned 1.

It can be shown using induction that the variable $v^a_{ki}$ (for $k>1$) represents the disjunction $\bigvee_{1\leq h\leq k-1}s^a_{hi} = s^a_{1i}\vee\dots \vee s^a_{(k-1)i}$.
From constraints \eqref{hipp-sat:clause1} and \eqref{hipp-sat:clause3} it follows that $v^a_{2i}$ is equivalent to $s^a_{1i}$.
Constraint \eqref{hipp-sat:clause3} also provides yields the proof of the inductive step.
Thus, constraint \eqref{hipp-sat:clause4} which sets $v^a_{(r+1)i}=1$ ensures that at least one of $s^a_{1i},\dots , s^a_{ri}$ is set to 1.
Furthermore, there can only be one $s^a_{ki}$ set to 1 due to the following reasoning.
Since $v^a_{ki}$ is equivalent to $\bigvee_{1\leq h\leq k-1}s^a_{hi} = s^a_{1i}\vee\dots \vee s^a_{(k-1)i}$ it follows that $v^a_{ki}$ being set 1 enforces all $v^a_{hi}$ to be set 1 for $h>k$.
Now assume there are two indices $h$ and $h'$ ($h<h'$) such that both $s^a_{hi}$ and $s^a_{h'i}$ are assigned true.
By the previous observation $v^a_{h'i}$ is also assigned true which violates constraint \eqref{hipp-sat:clause2}.

Next we formalize constraint (ii) stating that each genotype $g_i$ should be explained by its selected pair of haplotypes $(h_{k_1}, h_{k_2})$.
If $g_{ij}$ (the $j$'th entry of $g_i$) is 0, then the $j$'th entries of the chosen haplotypes should also be 0.
This is encoded by $s^a_{ki}\rightarrow\neg h_{kj}$ and $s^b_{ki}\rightarrow\neg h_{kj}$ for all $1\leq k \leq r$.
Similarly, if $g_{ij}$ is 1, then the $j$'th entries of the chosen haplotypes should also be 1, which is encoded by $s^a_{ki}\rightarrow h_{kj}$ and $s^b_{ki}\rightarrow h_{kj}$ for all $1\leq k \leq r$.
If $g_{ij}$ is 2, then the $j$'th entry of the haplotypes should differ.
Variables $g^a_{ij}$ and $g^b_{ij}$ are used to decide which haplotype should be assigned 0 and which should be assigned 1.
The clauses $(g^a_{ij}\vee g^b_{ij})\wedge(\neg g^a_{ij}\vee \neg g^b_{ij})$ enforce $g^a_{ij}$ and $g^b_{ij}$ to take on different values.
The constraints $s^a_{ki}\rightarrow (h_{kj}\leftrightarrow g^a_{ij})$ and $s^b_{ki}\rightarrow (h_{kj}\leftrightarrow g^b_{ij})$ then enforce the selected haplotypes to take on values dictated by $g^a_{ij}$ and $g^b_{ij}$.
\subsection{ILP formulations}
We explain here two different formulations of the HIPP problem, RTIP~\cite{DBLP:conf/cpm/Gusfield03} and PolyIP~\cite{DBLP:journals/tcbb/BrownH06}.
RTIP is an exponential-size encoding formulated and empirically evaluated by Gusfield in~\cite{DBLP:conf/cpm/Gusfield03}.
Similar polynomial-size encodings were simultaneously and independently formulated~\cite{DBLP:journals/tcbb/BrownH06, DBLP:conf/recomb/HalldorssonBELYI02, DBLP:journals/informs/LanciaPR04}.
Out of the polynomial size encodings we showcase PolyIP which is formulated by Brown and Harrower~\cite{DBLP:journals/tcbb/BrownH06}.

@@

The RTIP encoding contains variables representing the individual haplotype vectors as well as ones representing certain pairs of haplotype vectors (which correspond to a certain genotype).
Denote by $k_i$ the numbers of heterozygous loci (i.e., number of 2s in the vector) in genotype $g_i$.
Now there are $2^{k_i-1}$ distinct haplotype pairs corresponding to genotype $g_i$.
Denote by $L$ the number of distinct haplotypes appearing in the haplotype pairs of all $g_i\in\mathcal{G}$.
The encoding contains variables $y_{i,j}$ where $1\leq j \leq 2^{k-1}$ representing each of the possible haplotype pairs corresponding to $g_i$.
Similarly, each of the individual haplotype vectors appearing in the pairs is assigned a variable $x_l$.
Both the $y_{i,j}$ and $x_l$ variables are restricted to the range $[0,1]$, i.e. they may take only two distinct values, 0 or 1 (since only integer solutions are sought).
Setting $y_{i,j}$ to 1 indicates that $g_i$ is resolved using haplotype pair $p_j$ whereas 
setting $x_l$ to 1 indicates that haplotype $h_l$ is being used in some pair to resolve a genotype.
The constraint
$$ \text{for all  }i\in\{1,\dots , n\}:\quad \sum_{1\leq j \leq 2^{k_i-1}} y_{i,j} = 1 $$
ensures that each genotype is resolved by exactly one haplotype pair $p_j$.
Denote the haplotypes in pair $p_j$ by $h_{l(j,a)}$ and $h_{l(j,b)}$.
To ensure that the haplotypes of a used pair are marked used the encoding contains the following inequalities for all $i$ ($1\leq i \leq n$) and $j$ ($1\leq j \leq 2^{k_i-1}$).
\begin{eqnarray*}
y_{i,j} - x_{l(j,a)} & \leq & 0 \\
y_{i,j} - x_{l(j,b)} & \leq & 0 \\
\end{eqnarray*}
The objective of RTIP is to minize the number of used haplotypes, i.e., $\sum_{1\leq l \leq L} x_l$.
The entire encoding is presented in Figure~\ref{fig:enc-rtip}.

@@

The PolyIP encoding models the $2n$ haplotypes $h_i$, using $2nm$ binary variables $h_{i,j}$ where $1\leq i \leq 2n$ and $1\leq j \leq m$.
The haplotypes $h_{2i}$ and $h_{2i-1}$ are constrained by the encoding to correspond to genotype $g_i$ by including for each $i\in\{1, \dots , n\}$ and each $j\in\{1,\dots ,m\}$ the constraint
$$ h_{2i,j} + h_{2i-1,j} = g_{i,j}. $$
The encoding contains binary variables $d_{i,j}$ where $1\leq i,j\leq 2n$ to detect whether $h_i$ and $h_j$ are different.
The constraints 
$$ \text{for all }k\in\{1,\dots ,m\}:\quad  d_{i,j} \geq h_{i,k}-h_{j,k} $$
and
$$ \text{for all }k\in\{1,\dots ,m\}:\quad  d_{i,j} \geq h_{j,k}-h_{i,k} $$
ensure that $d_{i,j}$ is set to 1 whenever $h_i\neq h_j$.

@@

\begin{figure}
\centering
Minimize $\sum_{1\leq i \leq 2n} x_i$
\begin{align}
\text{for all  }i\in\{1,\dots , n\}:&\quad \sum_{1\leq j \leq 2^{k_i-1}} y_{i,j} = 1 \\
\end{align}
\caption{The PolyIP encoding.}
\label{fig:enc-polyip}
\end{figure}

>>>>

Denote by $\mathcal{G}=\{g_i\mid 1\leq i \leq n\}$ a set of $n$ genotype vectors of length $m$ and by $g_{i,j}$ the $j$'th entry of genotype vector $g_i$.
We construct a propositional formula $\phi (\mathcal{G}, r)$ which is satisfiable if and only if there exists a set of haplotypes $\mathcal{H}$ of size $r$ such that each genotype $g_i\in \mathcal{G}$ is explained by some pair fo haplotypes $(h_j, h_k)$ in $\mathcal{H}$.
To solve the HIPP problem the satisfiability of $\phi (\mathcal{G}, r)$ is checked for increasing values of $r$ until a value is found such that $\phi (\mathcal{G}, r)$ is satisfiable but $\phi (\mathcal{G}, r-1)$ is not.
This value of $r$ is then the minimum number of haplotypes required, and the actual haplotypes can be extracted from the satisfying assignment to $\phi (\mathcal{G}, r)$.

The formula $\phi (\mathcal{G}, r)$ considers a set of $r$ distinct haplotypes $h_k$ and encodes the assignment of a pair $(h_{k_1}, h_{k_2})$ to each $g_i$ using Boolean variables $s^a_{k_1,i}$, $s^b_{k_2,i}$ and $h_{k,j}$.
That is, variables $s^a_{k_1,i}$ and $s^b_{k_2,i}$ being set true indicates that $(h_{k_1}, h_{k_2})$ is the pair of haplotypes assigned to genotype $g_i$.
The variable $h_{k,j}$ is the $j$'th entry of haplotype $h_k$.
The formula then requires two types of constraints
\begin{enumerate}
\item[(i)] The selector variable $s^a_{k,i}$ should select a unique haplotype, i.e., for each genotype $g_i$ exactly one variable $s^a_{k,i}$ should be set true (and similarly for $s^b_{k,i}$). 
\item[(ii)] Each genotype $g_i$ should be explained by its selected pair of haplotypes $(h_{k_1}, h_{k_2})$. 
\end{enumerate}

Constraint (i) corresponds to equations $\sum_{1\leq k\leq r} s^a_{k,i} = 1$ and $\sum_{1\leq k\leq r} s^b_{k,i} = 1$ holding for all $i$ such that $1\leq i \leq n$.
This constraint is enforced in $\phi (\mathcal{R}, r)$ utilizing a propositional encoding of an \emph{1-bit adder circuit}.
The adder circuit uses variables $v^a_{ki}$ where $1\leq k \leq r+1$ and $1\leq i\leq n$ and the following constraints (the ones for $s^b_{k,i}$ are similar).
\begin{eqnarray}
\forall i\text{ such that } 1\leq i\leq n: & v^a_{1,i} \leftrightarrow s^a_{1,i}\label{hipp-sat:clause1}\\
\forall k, i\text{ such that } 1\leq k \leq r,\ 1\leq i\leq n: & \neg v^a_{k,i} \vee \neg s^a_{k,i}\label{hipp-sat:clause2}\\
\forall k, i\text{ such that } 1\leq k \leq r,\ 1\leq i\leq n: & v^a_{k+1,i} \leftrightarrow (s^a_{k,i}\vee v^a_{k,i} )\label{hipp-sat:clause3}\\
\forall i\text{ such that } 1\leq i\leq n: & v^a_{r+1,i}=1\label{hipp-sat:clause4}
\end{eqnarray}
Constraint \eqref{hipp-sat:clause1} enforces $v^a_{1,i}$ and $s^a_{1,i}$ to take on the same value in every satisfying assignment whereas constraint \eqref{hipp-sat:clause4} enforces variable $v^a_{r+1,i}$ to be evaluated to true.
Constraint \eqref{hipp-sat:clause2} prevents $v^a_{k,i}$ and $s^a_{k,i}$ from being simultaneously assigned true, 
and constraint \eqref{hipp-sat:clause3} sets $v^a_{k+1,i}$ to 1 whenever either $s^a_{k,i}$ or $v^a_{k,i}$ is assigned 1.

It can be shown using induction that the variable $v^a_{k,i}$ (for $k>1$) represents the disjunction $\bigvee_{1\leq h\leq k-1}s^a_{h,i} = s^a_{1,i}\vee\dots \vee s^a_{k-1,i}$.
From constraints \eqref{hipp-sat:clause1} and \eqref{hipp-sat:clause3} it follows that $v^a_{2,i}$ is equivalent to $s^a_{1,i}$.
Constraint \eqref{hipp-sat:clause3} also provides yields the proof of the inductive step.
Thus, constraint \eqref{hipp-sat:clause4} which sets $v^a_{r+1,i}=1$ ensures that at least one of $s^a_{1,i},\dots , s^a_{r,i}$ is set to 1.
Furthermore, there can only be one $s^a_{k,i}$ set to 1 due to the following reasoning.
Since $v^a_{k,i}$ is equivalent to $\bigvee_{1\leq h\leq k-1}s^a_{h,i} = s^a_{1,i}\vee\dots \vee s^a_{k-1,i}$ it follows that $v^a_{k,i}$ being set 1 enforces all $v^a_{h,i}$ to be set 1 for $h>k$.
Now assume there are two indices $h$ and $h'$ ($h<h'$) such that both $s^a_{h,i}$ and $s^a_{h',i}$ are assigned true.
By the previous observation $v^a_{h',i}$ is also assigned true which violates constraint \eqref{hipp-sat:clause2}.

Next we formalize constraint (ii) stating that each genotype $g_i$ should be explained by its selected pair of haplotypes $(h_{k_1}, h_{k_2})$.
If $g_{i,j}$ (the $j$'th entry of $g_i$) is 0, then the $j$'th entries of the chosen haplotypes should also be 0.
This is encoded by $s^a_{k,i}\rightarrow\neg h_{k,j}$ and $s^b_{k,i}\rightarrow\neg h_{k,j}$ for all $1\leq k \leq r$.
Similarly, if $g_{i,j}$ is 1, then the $j$'th entries of the chosen haplotypes should also be 1, which is encoded by $s^a_{k,i}\rightarrow h_{k,j}$ and $s^b_{k,i}\rightarrow h_{k,j}$ for all $1\leq k \leq r$.
If $g_{i,j}$ is 2, then the $j$'th entry of the haplotypes should differ.
Variables $g^a_{i,j}$ and $g^b_{i,j}$ are used to decide which haplotype should be assigned 0 and which should be assigned 1.
The clauses $(g^a_{i,j}\vee g^b_{i,j})\wedge(\neg g^a_{i,j}\vee \neg g^b_{i,j})$ enforce $g^a_{i,j}$ and $g^b_{i,j}$ to take on different values.
The constraints $s^a_{k,i}\rightarrow (h_{k,j}\leftrightarrow g^a_{i,j})$ and $s^b_{k,i}\rightarrow (h_{k,j}\leftrightarrow g^b_{i,j})$ then enforce the selected haplotypes to take on values dictated by $g^a_{i,j}$ and $g^b_{i,j}$.
\begin{figure}
\centering
\small
\begin{align}
\text{for all  }& i\in\{1,\dots , n\}:\nonumber\\
& \quad\quad\quad\quad \sum_{1\leq k \leq r} s^a_{k,i} = 1 \\
& \quad\quad\quad\quad \sum_{1\leq k \leq r} s^b_{k,i} = 1 \\
\text{for all  }& i\in\{1,\dots , n\},\ k\in\{1,\dots r\}\text{ and }\nonumber\\&\quad\quad j\in\{1,\dots ,m\} \text{ such that }g_{i,j}=0:\nonumber\\
& \quad\quad\quad\quad s^a_{k,i}\rightarrow \neg h_{k,j} \\
& \quad\quad\quad\quad s^b_{k,i}\rightarrow \neg h_{k,j} \\
\text{for all  }& i\in\{1,\dots , n\},\ k\in\{1,\dots r\}\text{ and }\nonumber\\&\quad\quad j\in\{1,\dots ,m\} \text{ such that }g_{i,j}=1:\nonumber\\
& \quad\quad\quad\quad s^a_{k,i}\rightarrow h_{k,j} \\
& \quad\quad\quad\quad s^b_{k,i}\rightarrow h_{k,j} \\
\text{for all  }& i\in\{1,\dots , n\},\ k\in\{1,\dots r\}\text{ and }\nonumber\\&\quad\quad j\in\{1,\dots ,m\} \text{ such that }g_{i,j}=2:\nonumber\\
& \quad\quad\quad\quad g^a_{i,j}\vee g^b_{i,j} \\
& \quad\quad\quad\quad \neg g^a_{i,j}\vee \neg g^b_{i,j} \\
& \quad\quad\quad\quad s^a_{k,i}\rightarrow (h_{k,j}\leftrightarrow g^a_{i,j}) \\
& \quad\quad\quad\quad s^b_{k,i}\rightarrow (h_{k,j}\leftrightarrow g^b_{i,j}) \\
\end{align}
\caption{The plain HIPP-SAT encoding.}
\label{fig:enc-hipp-sat}
\end{figure}
The encoding is presented in Figure~\ref{fig:enc-hipp-sat}.

% TODO: optimizations incl. symmetry-breaking
Lynce and Marques-Silva use two techniques to facilitate and speed up the use of their RTIP encoding~\cite{DBLP:conf/aaai/LynceM06}.
There are symmetries present in the plain SAT encoding, but fortunately these symmetries can be dispatched by utilizing symmetry-breaking predicates.
Since the usage of the encoding consists of calling $\phi (\mathcal{G}, r)$ with values of $r$ from some lower bound $lb$ up to $2n$ until the formula becomes satisfiable, the overall effort can be reduced by finding a large $lb$.
The computation of lower bounds of $r$ are thus useful in speeding up SAT-based haplotype inference.

Let $h$ and $h'$ be two concrete haplotype vectors.
Assume that the valuation of $h_{k_1}$ is $h$, the valuation of $h_{k_2}$ is $h'$ and that $s_{k_1, i}^a$ and  $s_{k_2, i}^b$ are set true.
This means that genotype $g_i$ is explained by haplotypes $h$ and $h'$.
Another equivalent solution to the SAT encoding can be constructed from this one by swapping the roles of $h_{k_1}$ and $h_{k_2}$ as well as those of $s_{k_1, i}^a$ and  $s_{k_2, i}^b$.
This is an example of a symmetry present in the plain formulation, 
which can be broken by enforcing the used haplotype to be in lexicographic order.
The ordering constraint should thus enforce that the for any valuation $v$ it holds that $v(h_1) < v(h_2) < \dots < v(h_r)$ where $v(h_i)$ is the binary string $v(h_{i,1})v(h_{i,2})\dots v(h_{i,n})$.
Frisch et al~\cite{DBLP:conf/cp/FrischHKMW02} present ways of encoding these types of ordering constraints.
Essentially constraints are added to the encoding that for each $i\in\{1,\dots ,r-1\}$ enforce that $v(h_i)$ is lexicographically less than $v(h_{i+1})$.

To find a lower bound $lb$ Lynce and Marques-Silva construct an \emph{incompatibility graph} of the genotypes~\cite{DBLP:conf/aaai/LynceM06}.
Two genotypes are \emph{incompatible} if there is a locus at which they are both homozygous but with different alleles.
The incompatibility graph has a node for each genotype and an edge between two nodes if the corresponding genotypes are incompatible.
Assume that a clique of size $k$ exists in the constructed incompatibility graph.
Then at least $2k - \sigma$ haplotypes are required where $\sigma$ is the number of genotypes in the clique without heterozygous loci.
The number $2k-\sigma$ can thus be taken as the lower bound $lb$ for $r$.
Since finding the largest clique in a graph is NP-hard it can be more beneficial to find a maximal clique using greedy heuristics.

\subsection{ILP formulations}
We explain here two different formulations of the HIPP problem, RTIP~\cite{DBLP:conf/cpm/Gusfield03} and PolyIP~\cite{DBLP:journals/tcbb/BrownH06}.
RTIP is an exponential-size encoding formulated and empirically evaluated by Gusfield in~\cite{DBLP:conf/cpm/Gusfield03}.
Similar polynomial-size encodings were simultaneously and independently formulated~\cite{DBLP:journals/tcbb/BrownH06, DBLP:conf/recomb/HalldorssonBELYI02, DBLP:journals/informs/LanciaPR04}.
Out of the polynomial size encodings we showcase PolyIP which is formulated by Brown and Harrower~\cite{DBLP:journals/tcbb/BrownH06}.

@@

The RTIP encoding contains variables representing the individual haplotype vectors as well as ones representing certain pairs of haplotype vectors (which correspond to a certain genotype).
Denote by $k_i$ the numbers of heterozygous loci (i.e., number of 2s in the vector) in genotype $g_i$.
Now there are $2^{k_i-1}$ distinct haplotype pairs corresponding to genotype $g_i$.
Denote by $L$ the number of distinct haplotypes appearing in the haplotype pairs of all $g_i\in\mathcal{G}$.
The encoding contains variables $y_{i,j}$ where $1\leq j \leq 2^{k_i-1}$ representing each of the possible haplotype pairs corresponding to $g_i$.
Similarly, each of the individual haplotype vectors appearing in the pairs is assigned a variable $x_l$.
Both the $y_{i,j}$ and $x_l$ variables are restricted to the range $[0,1]$, i.e. they may take only two distinct values, 0 or 1 (since only integer solutions are sought).
Setting $y_{i,j}$ to 1 indicates that $g_i$ is resolved using haplotype pair $p_j$ whereas 
setting $x_l$ to 1 indicates that haplotype $h_l$ is being used in some pair to resolve a genotype.
The constraint
$$ \text{for all  }i\in\{1,\dots , n\}:\quad \sum_{1\leq j \leq 2^{k_i-1}} y_{i,j} = 1 $$
ensures that each genotype is resolved by exactly one haplotype pair $p_j$.
Denote the haplotypes in pair $p_j$ by $h_{l(j,a)}$ and $h_{l(j,b)}$.
To ensure that the haplotypes of a used pair are marked used the encoding contains the following inequalities for all $i$ ($1\leq i \leq n$) and $j$ ($1\leq j \leq 2^{k_i-1}$).
\begin{eqnarray*}
y_{i,j} - x_{l(j,a)} & \leq & 0 \\
y_{i,j} - x_{l(j,b)} & \leq & 0 \\
\end{eqnarray*}
The objective of RTIP is to minize the number of used haplotypes, i.e., $\sum_{1\leq l \leq L} x_l$.
The entire encoding is presented in Figure~\ref{fig:enc-rtip}.

@@

The PolyIP encoding models the $2n$ haplotypes $h_i$, using $2nm$ binary variables $h_{i,j}$ where $1\leq i \leq 2n$ and $1\leq j \leq m$.
The haplotypes $h_{2i}$ and $h_{2i-1}$ are constrained by the encoding to correspond to genotype $g_i$ by including for each $i\in\{1, \dots , n\}$ and each $j\in\{1,\dots ,m\}$ the constraint
$$ h_{2i,j} + h_{2i-1,j} = g_{i,j}. $$
The encoding contains binary variables $d_{i,j}$ where $1\leq i<j\leq 2n$ to detect whether $h_i$ and $h_j$ are different.
The constraints 
$$ \text{for all }k\in\{1,\dots ,m\}:\quad  d_{i,j} \geq h_{i,k}-h_{j,k} $$
and
$$ \text{for all }k\in\{1,\dots ,m\}:\quad  d_{i,j} \geq h_{j,k}-h_{i,k} $$
ensure that $d_{i,j}$ is set to 1 whenever $h_i\neq h_j$.

@@

\begin{figure}
\centering
Minimize $\sum_{1\leq i \leq 2n} x_i$
\begin{align}
\text{for all  }i\in\{1,\dots , n\},\ j\in\{1,\dots ,m\}:&\quad h_{2i,j} + h_{2i-1,j} = g_{i,j} \\
\text{for all  }i, j\in\{1,\dots , 2n\},\text{ such that } i<j:&\quad d_{i,j} \geq h_{i,k}-h_{j,k} \\
&\quad d_{i,j} \geq h_{j,k}-h_{i,k} \\
\text{for all  }i\in\{1,\dots , 2n\}:&\quad x_{i} \geq 2 - i \sum_{j=1}^{i-1} d_{i,j} \\
\text{for all  }i,j\in\{1,\dots , 2n\},\ k\in\{1,\dots ,m\}:&\quad x_i\in\{0,1\} \\
&\quad h_{i,k}\in\{0,1\} \\
&\quad d_{i,j}\in\{0,1\}
\end{align}
\caption{The PolyIP encoding.}
\label{fig:enc-polyip}
\end{figure}

\section{Comparison}

\begin{itemize}
\item structure
\item Size
\item performance
\end{itemize}
